---
title: "Re-analysis of Crits-Christoph et al. (2021) using pipeline v2.4.0"
author: "Will Bradshaw"
date: 2024-10-17
format:
  html:
    code-fold: true
    code-tools: true
    code-link: true
    df-print: paged
    toc: true
    toc-depth: 2
    number-sections: true
    number-depth: 3
    crossref:
      fig-title: Figure
      fig-prefix: Figure
      chapters: true
editor: visual
title-block-banner: black
---

It's been quite a while since I last made a public notebook post, as I've been focusing on pipeline development and analysis of our increasing stream of private datasets. Recently, pipeline development has accelerated with the addition of Harmon Bhasin to the NAO computational team. Harmon and I recently finished a tranche of issue fixes and new features, and I'm writing this notebook as a way to both demo those features and test them before merging the new pipeline version to the main branch.

My initial analysis of the [Crits-Christoph dataset](https://doi.org/10.1128%2FmBio.02703-20) was quite some time ago now ([1](https://data.securebio.org/wills-public-notebook/notebooks/2024-02-04_crits-christoph-1.html),[2](https://data.securebio.org/wills-public-notebook/notebooks/2024-02-08_crits-christoph-2.html),[3](https://data.securebio.org/wills-public-notebook/notebooks/2024-02-15_crits-christoph-3.html)), back when I was still figuring out the basic functionality of the pipeline. We've come quite a long way since then, so this is a good opportunity to apply a more streamlined and well-tested analysis process to a very early dataset.

As a reminder, the Crits-Christoph data comprises 18 samples, all of which were collected as 24-hour composite samples of raw sewage collected from wastewater interceptor lines in four different locations in the SF Bay Area (Berkeley, Oakland, San Francisco, and Marin). Several different extraction protocols were used; for simplicity, I won't be including this as a variable in this analysis, since the number of samples is too small to draw strong conclusions about inter-protocol differences.

Seven of the samples (the "unenriched" fraction) underwent bacterial ribodepletion followed by untargeted metagenomic RNA-sequencing. The other 11 samples (the “enriched” fraction) underwent panel enrichment for respiratory viruses using the Illumina Respiratory Virus Oligo Panel, and no ribodepletion. In total, the unenriched samples comprise roughly 298M read pairs (mean 42.5M per sample) while the enriched samples comprise roughly 9M (mean 0.8M per sample).

```{r}
#| label: load-packages
#| include: false

# Import libraries
library(Biostrings)
library(tidyverse)
library(cowplot)
library(patchwork)
library(fastqcr)
library(RColorBrewer)
library(ggpubr)
library(ggbeeswarm)

# Import auxiliary scripts
source("../scripts/aux_plot-theme.R")
source("../scripts/aux_functions.R")

# Scales and palettes
scale_fill_loc <- purrr::partial(scale_fill_brewer, name="Location",
                                 palette="Dark2")
scale_color_loc <- purrr::partial(scale_color_brewer, name="Location",
                                 palette="Dark2")
scale_shape_loc <- purrr::partial(scale_shape_discrete, name = "Location")

```

# Cost profiling

As for the previous datasets I've run through this version of the pipeline, I generated data on computing resources consumed using `nextflow log` and AWS Cost Explorer.

```{r}
#| label: define-log-data
#| edit-manually: true

# Set input dir
data_dir <- "../data/2024-10-17_crits-christoph-2-4-0"
log_dir <- file.path(data_dir, "logging")

# Profiling parameters
tab_log <- tibble(
  dataset = "Crits-Christoph et al. 2021",
  read_pairs = 306407694,
  n_samples = 18
)

# Import trace
trace_path <- file.path(log_dir, "trace-2024-10-17_16-27-24.txt")
trace <- lapply(trace_path, read_tsv, show_col_types = FALSE) %>% bind_rows %>%
  mutate(realtime_s = parse_durations(realtime),
         cpu_hours = realtime_s * cpus / 3600,
         workflow = process %>% str_split(":") %>% sapply(dplyr::first),
         job = process %>% str_split(":") %>% sapply(dplyr::last),
         subworkflow = process %>% middle(":"),
         dataset = "Crits-Christoph et al. 2021"
         )


```

```{r}
#| label: process-trace-data
#| edit-manually: false

# Parse trace data
trace_cpuh_summary <- trace %>% group_by(dataset, job) %>%
  summarize(cpu_hours = sum(cpu_hours), .groups = "drop_last") %>%
  mutate(p_cpu_hours = cpu_hours/sum(cpu_hours))
cpuh_total <- trace_cpuh_summary %>% summarize(cpu_hours = sum(cpu_hours))

#==============================================================================
# Aggregate data for plotting
#==============================================================================

# Group processes across subworkflows
jobs_major <- trace_cpuh_summary %>% group_by(job) %>%
  filter(p_cpu_hours == max(p_cpu_hours)) %>%
  arrange(desc(p_cpu_hours)) %>% ungroup %>%
  head(8) %>% pull(job)
trace_cpuh_display <- trace_cpuh_summary %>%
  mutate(minor = !(job %in% jobs_major),
         job_display = ifelse(minor, "Other", job)) %>%
  group_by(dataset, job_display, minor) %>%
  summarize(cpu_hours = sum(cpu_hours), .groups = "drop") %>%
  arrange(minor, desc(cpu_hours)) %>%
  mutate(job_display = fct_inorder(job_display))

# Break down log files into subworkflows, segregating BBDUK from the rest of the HV workflow
cat_levels <- c("Preprocessing", "Taxonomic profiling",
                "HV identification (BBDuk)", 
                "HV identification (other)", "Other")
trace_cpuh_categories <- trace %>%
  mutate(category = case_when(
    grepl("^RUN:RAW", process) ~ "Preprocessing",
    grepl("^RUN:CLEAN", process) ~ "Preprocessing",
    grepl("^RUN:PROFILE", process) ~ "Taxonomic profiling",
    grepl("^RUN:HV:BBDUK", process) ~ "HV identification (BBDuk)",
    grepl("^RUN:HV", process) ~ "HV identification (other)",
    TRUE ~ "Other"
  )) %>%
  group_by(dataset, category) %>% summarize(cpu_hours = sum(cpu_hours), .groups = "drop") %>%
  mutate(p_cpu_hours = cpu_hours / sum(cpu_hours),
         category = factor(category, levels = cat_levels))
```

In total, processing the Crits-Christoph data consumed roughly 99 CPU-hours. While I don't have AWS costs for this run, at our usual cost per CPU-hour, I estimate this cost roughly \$5-8.

Here's the breakdown of CPU-hours by process and subworkflow:

```{r}
#| label: fig-job-compute
#| edit-manually: false
#| fig-cap: "Pie charts of CPU-hour composition of v2.3 pipeline across processes."

# Group by run group rather than dataset
trace_cpuh_display_agg <- trace_cpuh_display %>% 
  group_by(job_display, minor) %>% 
  summarize(cpu_hours = sum(cpu_hours), .groups = "drop")

# Plot
g_log_pie <- ggplot(trace_cpuh_display_agg, 
                    aes(x=1, y=cpu_hours, fill=job_display)) +
  geom_col(width=1, position="fill") +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette="Set1", name="Process") +
  theme_void()
g_log_pie
```

```{r}
#| label: fig-subprocess-compute
#| edit-manually: false
#| fig-cap: "Pie charts of CPU-hour composition of v2.3 pipeline across process categories"

# Group by run group rather than dataset
trace_cpuh_categories_agg <- trace_cpuh_categories %>% 
  group_by(category) %>% 
  summarize(cpu_hours = sum(cpu_hours), .groups = "drop")

# Plot
g_cat_pie <- ggplot(trace_cpuh_categories_agg,
                    aes(x=1, y=cpu_hours, fill=category)) +
  geom_col(width=1, position="fill") +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette="Set3", name="Process category") +
  theme_void()
g_cat_pie
```

# Raw data & preprocessing

## Read counts

```{r}
#| label: prepare-libraries
#| edit-manually: true

# Define directory structure
output_dirs <- data_dir
input_dirs <- file.path(output_dirs, "input")
results_dirs <- file.path(output_dirs, "results")

# Get metadata paths
metadata_paths <- file.path(output_dirs, "metadata.csv")

# Import metadata
metadata_raw <- lapply(metadata_paths, read_csv, show_col_types = FALSE) %>%
  bind_rows

# Process metadata
metadata <- metadata_raw %>%
  rename(sample_alias = sample, sample=library, date=collection_date) %>%
  mutate(enrichment = str_to_title(enrichment)) %>%
  arrange(date) %>%
  mutate(date = fct_inorder(as.character(date)))

# Define grouping columns
grouping_cols <- c("date", "location", "enrichment")
grouping_cols_agg <- c("enrichment")
```

```{r}
#| warning: false
#| label: read-qc-data
#| edit-manually: false

# Data input paths
qc_dirs <- file.path(results_dirs, "qc")
hv_dirs <- file.path(results_dirs, "hv")
basic_stats_paths <- file.path(qc_dirs, "qc_basic_stats.tsv.gz")
adapter_stats_paths <- file.path(qc_dirs, "qc_adapter_stats.tsv.gz")
quality_base_stats_paths <- file.path(qc_dirs, "qc_quality_base_stats.tsv.gz")
quality_seq_stats_paths <- file.path(qc_dirs, "qc_quality_sequence_stats.tsv.gz")

# Import QC data
stages <- c("raw_concat", "cleaned")
basic_stats <- lapply(basic_stats_paths, read_tsv, show_col_types = FALSE) %>%
  bind_rows %>%
  inner_join(metadata, by="sample") %>% arrange(sample) %>%
  mutate(stage = factor(stage, levels = stages),
         sample = fct_inorder(sample))
adapter_stats <- lapply(adapter_stats_paths, read_tsv, show_col_types = FALSE) %>%
  bind_rows %>%
  inner_join(metadata, by="sample") %>% arrange(sample) %>%
  mutate(stage = factor(stage, levels = stages),
         read_pair = fct_inorder(as.character(read_pair)))
quality_base_stats <- lapply(quality_base_stats_paths, read_tsv, show_col_types = FALSE) %>%
  bind_rows %>%
  inner_join(metadata, by="sample") %>% arrange(sample) %>%
  mutate(stage = factor(stage, levels = stages),
         read_pair = fct_inorder(as.character(read_pair)))
quality_seq_stats <- lapply(quality_seq_stats_paths, read_tsv, show_col_types = FALSE) %>%
  bind_rows %>%
  inner_join(metadata, by="sample") %>% arrange(sample) %>%
  mutate(stage = factor(stage, levels = stages),
         read_pair = fct_inorder(as.character(read_pair)))

basic_stats_raw <- basic_stats %>% filter(stage == "raw_concat")
```

```{r}
#| label: summarize-qc-data
#| edit-manually: true

get_read_counts <- function(tab) {
  summarize(tab, rmin = min(n_read_pairs), rmax=max(n_read_pairs),
            rmean=mean(n_read_pairs), 
            rtot = sum(n_read_pairs),
            btot = sum(n_bases_approx),
            .groups = "drop")
}

read_counts_total <- basic_stats_raw %>% ungroup %>% get_read_counts
read_counts_grp <- basic_stats_raw %>% group_by(across(all_of(grouping_cols))) %>%
  get_read_counts
read_counts_agg <- basic_stats_raw %>% 
  group_by(across(all_of(grouping_cols_agg))) %>%
  get_read_counts
```

The 18 libraries in the Crits-Christoph dataset comprised roughly 306M read pairs, corresponding to roughly 46 gigabases of sequence data. The great majority of this data (\~97%) was generated from unenriched libraries.

```{r}
#| fig-width: 9
#| warning: false
#| label: fig-basic-stats
#| edit-manually: true
#| fig-cap: "Read and base counts for different samples."

# Prepare data
basic_stats_raw_metrics <- basic_stats_raw %>%
  mutate(n_read_pairs = replace_na(n_read_pairs, 0),
         n_bases_approx = replace_na(n_bases_approx, 0)) %>%
  select(sample, all_of(grouping_cols),
         `# Read pairs` = n_read_pairs,
         `Total base pairs\n(approx)` = n_bases_approx) %>%
  pivot_longer(-(sample:last(grouping_cols)), names_to = "metric", values_to = "value") %>%
  mutate(metric = fct_inorder(metric))

# Set up plot templates
g_basic <- ggplot(basic_stats_raw_metrics, 
                  aes(x=enrichment, y=value, fill=location, group=sample)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  scale_x_discrete() +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_loc() +
  expand_limits(y=c(0,100)) +
  facet_grid(metric~., scales = "free", space="free_x", switch="y") +
  theme_tilt + theme(
    axis.title.y = element_blank(),
    strip.text.y = element_text(face="plain")
  )
g_basic
```

On average, read preprocessing with FASTP resulted in the loss of roughly 2.1% of reads in unenriched libraries and 6.8% in enriched libraries. A minority of enriched libraries showed even higher losses:

```{r}
#| label: count-reads
#| edit-manually: false

# Count read losses
n_reads_rel <- basic_stats %>% 
  select(sample, all_of(grouping_cols), 
         stage, percent_duplicates, n_read_pairs) %>%
  group_by(sample, across(all_of(grouping_cols_agg))) %>% 
  arrange(sample, stage) %>%
  mutate(p_reads_retained = n_read_pairs / lag(n_read_pairs),
         p_reads_lost = 1 - p_reads_retained)
```

```{r}
#| label: fig-preproc-figures
#| warning: false
#| fig-height: 5
#| edit-manually: true
#| fig-cap: "Relative read losses during cleaning with FASTP."

# Plot relative read losses during preprocessing
g_reads_rel <- ggplot(n_reads_rel %>% filter(stage == "cleaned"), 
                      aes(x=enrichment, y=p_reads_lost, fill=location, group=sample)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  scale_y_continuous("% Total Reads Lost", expand=c(0,0), 
                     labels = function(x) x*100,
                     breaks=seq(0,0.2,0.01)) +
  scale_fill_loc() +
  theme_tilt
g_reads_rel
```

## Adapters

Adapter levels in the raw reads were moderate, and were very effectively removed by preprocessing with FASTP:

```{r}
#| label: fig-adapters
#| fig-cap: "FASTQC-measured adapter contamination in raw and FASTP-cleaned reads."
#| edit-manually: true
#| fig-width: 10

g_qual <- ggplot(mapping=aes(color=location, linetype=read_pair, 
                         group=interaction(sample,read_pair))) + 
  scale_color_loc() + scale_linetype_discrete(name = "Read Pair") +
  guides(color=guide_legend(nrow=2,byrow=TRUE),
         linetype = guide_legend(nrow=2,byrow=TRUE)) +
  theme_base

# Visualize adapters
g_adapters <- g_qual + 
  geom_line(aes(x=position, y=pc_adapters), data=adapter_stats) +
  scale_y_continuous(name="% Adapters", limits=c(0,NA),
                     breaks = seq(0,1000,10), expand=c(0,0)) +
  scale_x_continuous(name="Position along read", limits=c(0,NA),
                     breaks=seq(0,140,20), expand=c(0,0)) +
  facet_grid(stage~adapter)
g_adapters

```

## Quality

Qualities were high across deliveries, both at the per-base and per-read level:

```{r}
#| label: fig-quality-base
#| fig-cap: "FASTQC-measured per-base quality scores."
#| fig-width: 8
#| edit-manually: true

g_quality_base <- g_qual +
  geom_hline(yintercept=25, linetype="dashed", color="red") +
  geom_hline(yintercept=30, linetype="dashed", color="red") +
  geom_line(aes(x=position, y=mean_phred_score), data=quality_base_stats) +
  scale_y_continuous(name="Mean Phred score", expand=c(0,0), limits=c(10,45)) +
  scale_x_continuous(name="Position", limits=c(0,NA),
                     breaks=seq(0,140,20), expand=c(0,0)) +
  facet_grid(stage~.)
g_quality_base
```

```{r}
#| label: fig-quality-read
#| fig-cap: "FASTQC-measured per-read quality scores."
#| edit-manually: false
#| fig-width: 8

g_quality_seq <- g_qual +
  geom_vline(xintercept=25, linetype="dashed", color="red") +
  geom_vline(xintercept=30, linetype="dashed", color="red") +
  geom_line(aes(x=mean_phred_score, y=n_sequences), data=quality_seq_stats) +
  scale_x_continuous(name="Mean Phred score", expand=c(0,0)) +
  scale_y_continuous(name="# Sequences", expand=c(0,0)) +
  facet_grid(stage~., scales = "free_y")
g_quality_seq
```

## GC content

Unenriched samples show somewhat lower average GC content than enriched samples, with the former showing a mean GC content of 43.4% and the latter 47.9%:

```{r}
#| label: fig-gc-qc
#| fig-cap: "Distribution of GC content across samples."

# Get raw GC content
gc_content <- basic_stats_raw %>%
  group_by(across(all_of(grouping_cols))) %>%
  summarize(mean_percent_gc = mean(percent_gc),
            .groups = "drop")

# Plot
g_gc_st <- ggplot(gc_content, aes(x=enrichment, y=mean_percent_gc)) +
  geom_boxplot() +
  geom_quasirandom(alpha=0.7, shape=16) +
  scale_y_continuous(name="Mean GC content (% bases)") +
  coord_flip() +
  guides(color=guide_legend(nrow=2)) +
  theme_base + theme(axis.title.y = element_blank())
g_gc_st
```

# Taxonomic profiling

## Domain-level classification

As usual for wastewater, samples across all deliveries were dominated by bacterial and unassigned sequences:

```{r}
#| label: process-taxonomy-domains
#| edit-manually: false

# Import Bracken data
bracken_paths <- file.path(results_dirs, "taxonomy/bracken_reports_merged.tsv.gz")
bracken_tab <- lapply(bracken_paths, read_tsv, show_col_types = FALSE) %>% bind_rows %>%
  inner_join(metadata, by="sample") %>% 
  mutate(ribosomal_label = ifelse(ribosomal, "Ribosomal", "Non-ribosomal"))

# Import Kraken data
kraken_paths <- file.path(results_dirs, "taxonomy/kraken_reports_merged.tsv.gz")
kraken_tab <- lapply(kraken_paths, read_tsv, show_col_types = FALSE) %>% bind_rows %>%
  inner_join(metadata, by="sample") %>% 
  mutate(ribosomal_label = ifelse(ribosomal, "Ribosomal", "Non-ribosomal"))

# Extract taxon reads from Bracken and unassigned from Kraken
class_tab <- bracken_tab %>%
  select(all_of(grouping_cols),
         n_reads=new_est_reads, name, ribosomal_label)
unclass_tab <- kraken_tab %>% filter(taxid == 0) %>%
  select(all_of(grouping_cols),
         n_reads=n_reads_clade, name, ribosomal_label) %>%
  mutate(name=str_to_title(name))
taxa_tab_raw <- bind_rows(class_tab, unclass_tab)
taxon_levels <- expand_grid(
  taxon = c("Unclassified", "Bacteria", "Archaea", "Eukaryota", "Viruses"),
  ribo  = c("Ribosomal", "Non-ribosomal") 
) %>% mutate(label = paste0(taxon, " (", ribo, ")")) %>% pull(label)
taxa_tab <- mutate(taxa_tab_raw, 
                   label = paste0(name, " (", ribosomal_label, ")"),
                   label = factor(label, levels = taxon_levels))
taxa_tab_display <- taxa_tab
```

```{r}
#| label: fig-taxonomy-domains
#| fig-height: 6
#| fig-width: 6
#| edit-manually: true
#| fig-cap: "Domain-level composition of samples, as measured with Kraken2 on 1M reads per sample."

# Plot
g_bracken <- ggplot(mapping = aes(x=location, y=n_reads, fill=label)) +
  scale_y_continuous(name="% Reads", label = function(y) y*100) +
  facet_grid(.~enrichment, scales = "free_x", space = "free_x",
             labeller = label_wrap_gen(width = 10)) +
  guides(fill = guide_legend(ncol=3)) +
  theme_tilt
g_bracken_1 <- g_bracken +
    geom_col(data = taxa_tab_display, position = "fill", width=1) +
    scale_fill_brewer(palette="Set3", name="Taxon") +
    ggtitle("Taxonomic composition (all reads)")
g_bracken_1
```

## Ribosomal content

The above plot makes it clear that, consistent with previous analysis of this dataset, the unenriched samples (which underwent ribodepletion) show much lower ribosomal content than the enriched samples (which did not), despite the latter having undergone viral panel enrichment. Quantitatively, enriched samples contained 63.4% ribosomal sequences, while unenriched samples contained only 5.5%.

```{r}
#| label: ribo-summary
p_ribo_sample <- taxa_tab %>%
  group_by(across(all_of(grouping_cols)), ribosomal_label) %>%
  summarize(n_reads = sum(n_reads), .groups = "drop_last") %>%
  mutate(p_reads = n_reads/sum(n_reads)) %>%
  filter(ribosomal_label == "Ribosomal")
p_ribo_agg <- taxa_tab %>%
  group_by(across(all_of(grouping_cols_agg)), ribosomal_label) %>%
  summarize(n_reads = sum(n_reads), .groups = "drop_last") %>%
  mutate(p_reads = n_reads/sum(n_reads)) %>%
  filter(ribosomal_label == "Ribosomal")
```

```{r}
#| label: fig-ribo-box
#| fig-cap: "Boxplots of ribosomal content by enrichment status."
g_ribo_box <- ggplot(p_ribo_sample, aes(x=enrichment, y=p_reads)) +
  geom_boxplot() +
  geom_quasirandom(alpha=0.7, shape=16, mapping=aes(color=location)) +
  coord_flip() +
  scale_color_loc() +
  scale_y_continuous(name="Ribosomal content (% reads)",
                     breaks=seq(0,1,0.2),
                     labels=function(y)y*100) +
  guides(color=guide_legend(nrow=2)) +
  theme_base + theme(axis.title.y = element_blank())
g_ribo_box
```

## Total viral content

```{r}
#| label: process-viral-content
#| edit-manually: false

# Viral content across all reads
p_reads_viral_all <- taxa_tab %>% mutate(viral = name == "Viruses") %>%
  group_by(across(all_of(grouping_cols)), viral) %>%
  summarize(n_reads = sum(n_reads), .groups = "drop_last") %>% 
  mutate(p_reads = n_reads/sum(n_reads)) %>%
  filter(viral)

# Aggregate
p_reads_viral_all_agg <- taxa_tab %>% mutate(viral = name == "Viruses") %>%
  group_by(across(all_of(grouping_cols_agg)), viral) %>%
  summarize(n_reads = sum(n_reads), .groups = "drop_last") %>% 
  mutate(p_reads = n_reads/sum(n_reads)) %>%
  filter(viral)
```

Total viral content in the Kraken data was 4.0% in the enriched samples and 0.77% in the unenriched samples, demonstrating the efficacy of panel enrichment:

```{r}
#| label: fig-viral-content
#| edit-manually: true
#| warning: false
#| fig-cap: "Fraction of viral reads across samples, as measured with Kraken2 on 1M reads per sample."

g_viral_box <- ggplot(p_reads_viral_all, aes(x=enrichment, y=p_reads)) +
  geom_boxplot() +
  geom_quasirandom(alpha=0.7, shape=16, mapping=aes(color=location)) +
  coord_flip() +
  scale_color_loc() +
  scale_y_log10(name="Viral read fraction") +
  guides(color=guide_legend(nrow=2)) +
  theme_base + theme(axis.title.y = element_blank())
g_viral_box
```

## Taxonomic composition of viruses

```{r}
#| label: prepare-viral-taxonomy-plotting
#| edit-manually: true

# Set up base plot
g_comp_base <- ggplot(mapping = aes(x=location, y=p_reads, fill=label)) +
  facet_grid(.~enrichment, scales = "free_x", space = "free_x", labeller = label_wrap_gen(width = 10)) +
  guides(fill = guide_legend(ncol=3)) +
  theme_tilt

# Specify palette
palette_viral <- c(brewer.pal(12, "Set3"), brewer.pal(8, "Dark2"))
scale_fill_viral <- purrr::partial(scale_fill_manual, values = palette_viral)

# Set up composition scale
scale_y_composition <- purrr::partial(scale_y_continuous, limits = c(0,1.01),
                                      breaks = seq(0,1,0.2), expand = c(0,0),
                                      labels = function(y) y*100)

# Set up geom
geom_composition <- purrr::partial(geom_col, position = "fill", width = 1)

# Specify major family threshold
major_threshold <- 0.01
```

```{r}
#| label: extract-viral-taxa
#| edit-manually: false

# Get viral taxonomy
viral_taxa_path <- file.path(data_dir, "total-virus-db.tsv.gz")
viral_taxa <- read_tsv(viral_taxa_path, show_col_types = FALSE)

# Prepare viral Kraken tab
kraken_tab_viral_raw <- filter(kraken_tab, taxid %in% viral_taxa$taxid)
kraken_tab_viral_sum_noribo <- kraken_tab_viral_raw %>% filter(!ribosomal) %>%
  group_by(taxid, name, rank, across(all_of(grouping_cols))) %>%
  summarize(n_reads_clade = sum(n_reads_clade),
            n_reads_direct = sum(n_reads_direct),
            n_minimizers_total = sum(n_minimizers_total),
            n_minimizers_distinct = sum(n_minimizers_distinct),
            n_reads_clade_ribosomal = sum(n_reads_clade[ribosomal]),
            .groups = "drop") %>%
  mutate(p_reads_clade_ribosomal = n_reads_clade_ribosomal/n_reads_clade)
kraken_tab_viral_total_noribo <- kraken_tab_viral_sum_noribo %>%
  filter(taxid == 10239) %>%
  select(all_of(grouping_cols), n_reads_viral = n_reads_clade)
kraken_tab_viral_noribo <- kraken_tab_viral_sum_noribo %>%
  inner_join(kraken_tab_viral_total_noribo, by = grouping_cols) %>%
  mutate(p_reads_viral = n_reads_clade/n_reads_viral)
kraken_tab_viral_cleaned_noribo <- kraken_tab_viral_noribo %>%
  select(name, taxid, rank, all_of(grouping_cols), 
         n_reads_clade, n_reads_viral, p_reads_viral, p_reads_clade_ribosomal)

# Subset to specific taxonomic ranks
viral_classes_noribo <- kraken_tab_viral_cleaned_noribo %>% filter(rank == "C")
viral_families_noribo <- kraken_tab_viral_cleaned_noribo %>% filter(rank == "F")
```

As is normal for wastewater samples, the unenriched samples are dominated by *Alsuviricetes* (a class of eukaryotic viruses which includes *Virgaviridae*), *Leviviricetes* (a class of RNA phages), and *Caudoviricetes* (tailed DNA phages). *Alsuviricetes* and *Caudoviricetes* were also prominent in enriched samples; however, two other virus classes, *Pisoniviricetes* (which includes *Picornaviridae* and *Coronaviridae*, among other animal viruses) and *Quintoviricetes* (parvoviruses) are also prominent. Both of these additional classes include prominent human pathogens that are included in the Illumina Respiratory Virus Panel.

```{r}
#| label: fig-viral-class-composition-noribo
#| fig-height: 6
#| fig-width: 6
#| edit-manually: false
#| fig-cap: "Class composition of Kraken2-identified viral reads across samples, excluding putative ribosomal reads."

# Identify major viral classes and collapse others
viral_classes_major_noribo <- viral_classes_noribo %>%
  group_by(name, taxid) %>%
  filter(max(p_reads_viral) >= major_threshold) %>% ungroup
viral_classes_minor_noribo <- viral_classes_major_noribo %>%
  group_by(across(all_of(grouping_cols)), n_reads_viral) %>%
  summarize(n_reads_clade = n_reads_viral[1] - sum(n_reads_clade),
            p_reads_viral = 1 - sum(p_reads_viral), .groups = "drop") %>%
  mutate(name = "Other", taxid = NA, rank = "F")
viral_classes_levels_noribo <- viral_classes_major_noribo %>% pull(name) %>% sort %>% 
  unique %>% append("Other")
viral_classes_out_noribo <- bind_rows(viral_classes_major_noribo, viral_classes_minor_noribo) %>%
  mutate(name = factor(name, levels = viral_classes_levels_noribo))

# Prepare data for plotting
viral_classes_display_noribo <- viral_classes_out_noribo %>%
  rename(p_reads = p_reads_viral, label = name)

# Plot
g_classes_noribo <- g_comp_base + 
  geom_composition(data=viral_classes_display_noribo) +
  ggtitle("Viral class composition (non-ribosomal reads only)") +
  scale_y_composition(name="% Viral Reads") +
  scale_fill_viral(name="Viral Class")
g_classes_noribo
```

As usual, *Alsuviricetes* is overwhelmingly dominated by *Virgaviridae*, with a small minority of other plant viruses:

```{r}
#| label: fig-alsuviricetes-families
#| fig-cap: "Family composition of Kraken2-identified *Alsuviricetes* reads, excluding putative ribosomal reads."
#| fig-height: 6
#| fig-width: 6

# Get reads for major eukaryotic viral classes
euk_classes <- c("Alsuviricetes", "Pisoniviricetes", "Quintoviricetes")
euk_taxids <- sapply(euk_classes, function(x) viral_taxa %>% filter(name == x) %>% pull(taxid))
euk_taxid_desc <- sapply(euk_taxids, add_descendent_taxids, reference=viral_taxa)
euk_families <- lapply(euk_classes, function(x) viral_families_noribo %>% 
                         filter(taxid %in% euk_taxid_desc[[x]]) %>% mutate(class=x)) %>%
  bind_rows %>% group_by(across(all_of(grouping_cols)), class) %>%
  mutate(n_reads_class = sum(n_reads_clade),
         p_reads_class = n_reads_clade/n_reads_class)
euk_families_display <- lapply(euk_classes, function(x)
  euk_families %>% filter(class==x) %>% rename(p_reads = p_reads_class, label = name)) %>%
  setNames(euk_classes)

# Plot
g_euk_families <- lapply(euk_classes, function(x)
  g_comp_base + geom_composition(data=euk_families_display[[x]]) +
    scale_y_composition(name=paste0("% ", x, " Reads")) +
    scale_fill_viral(name="Viral Family")) %>% setNames(euk_classes)
g_euk_families[["Alsuviricetes"]]
```

*Pisoniviricetes* reads in enriched samples, meanwhile, are overwhelmingly made up of coronaviruses:

```{r}
#| label: fig-stelpaviricetes-families
#| fig-cap: "Family composition of Kraken2-identified *Pisoniviricetes* reads across samples, excluding putative ribosomal reads."
#| fig-height: 6
#| fig-width: 6

g_euk_families[["Pisoniviricetes"]]
```

# Human-infecting virus reads

```{r}
#| label: prepare-hv
#| edit-manually: false

# Import and format reads
hv_reads_paths <- file.path(hv_dirs, "hv_hits_putative_collapsed.tsv.gz")
mrg_hv <- lapply(hv_reads_paths, read_tsv, show_col_types = FALSE,
                   col_types = cols(taxid_all = "c", assigned_taxid_all = "c")) %>%
  bind_rows %>%
  inner_join(metadata, by="sample") %>%
  mutate(kraken_label = ifelse(assigned_hv, "Kraken2 HV assignment",
                               "No Kraken2 assignment"),
         adj_score_max = pmax(adj_score_fwd, adj_score_rev),
         highscore = adj_score_max >= 20)

# Raise to the class level and exclude phages
phage_taxids <- c("2731619", "2732413", "2732411") # Caudoviricetes, Malgrantaviricetes, Faserviricetes
mrg_hv_named <- mrg_hv %>% left_join(viral_taxa, by="taxid") 
mrg_hv_class <- raise_rank(mrg_hv_named, viral_taxa, "class")
mrg_hv_class_filtered <- filter(mrg_hv_class, !taxid %in% phage_taxids)
mrg_hv_filtered <- mrg_hv %>% filter(seq_id %in% mrg_hv_class_filtered$seq_id)

# Get read counts and fractions
read_counts_raw <- select(basic_stats_raw, sample, n_reads_raw = n_read_pairs)
read_counts_hv <- count(mrg_hv_filtered, sample, name="n_reads_hv")
read_counts <- left_join(read_counts_raw, read_counts_hv, by="sample") %>%
  mutate(n_reads_hv = replace_na(n_reads_hv, 0)) %>%
  inner_join(metadata, by="sample") %>%
  select(sample, all_of(grouping_cols), n_reads_raw, n_reads_hv) %>%
  mutate(n_samples = 1, p_reads_total = n_reads_hv/n_reads_raw)

# Aggregate read counts
read_counts_agg <- read_counts %>%
  group_by(across(all_of(grouping_cols_agg))) %>%
  summarize(n_reads_raw = sum(n_reads_raw),
            n_reads_hv = sum(n_reads_hv), 
            n_samples = sum(n_samples), .groups="drop") %>%
  mutate(p_reads_total = n_reads_hv/n_reads_raw)
```

## Overall relative abundance

Next, I calculated the number of human-infecting virus (HV) reads as a fraction of total raw reads:

```{r}
#| label: count-hv-reads
#| warning: false
#| edit-manually: false

# Get read counts and fractions
read_counts_raw <- select(basic_stats_raw, sample, n_reads_raw = n_read_pairs)
read_counts_hv <- count(mrg_hv_filtered, sample, name="n_reads_hv")
read_counts <- left_join(read_counts_raw, read_counts_hv, by="sample") %>%
  mutate(n_reads_hv = replace_na(n_reads_hv, 0)) %>%
  inner_join(metadata, by="sample") %>%
  select(all_of(grouping_cols), n_reads_raw, n_reads_hv) %>%
  group_by(across(all_of(grouping_cols))) %>%
  summarize(n_reads_raw = sum(n_reads_raw),
            n_reads_hv = sum(n_reads_hv), .groups = "drop") %>%
  mutate(n_samples = 1, p_reads_total = n_reads_hv/n_reads_raw)

# Aggregate read counts
read_counts_agg <- read_counts %>%
  group_by(across(all_of(grouping_cols_agg))) %>%
  summarize(n_reads_raw = sum(n_reads_raw),
            n_reads_hv = sum(n_reads_hv), 
            n_samples = sum(n_samples), .groups="drop") %>%
  mutate(p_reads_total = n_reads_hv/n_reads_raw)
```

```{r}
#| label: fig-hv-content
#| edit-manually: true
#| warning: false
#| fig-cap: "Unique human-infecting virus reads as a fraction of all raw input reads, for enriched and unenriched samples."

g_hv_box <- ggplot(read_counts, aes(x=enrichment, y=p_reads_total)) +
  geom_boxplot() +
  geom_quasirandom(alpha=0.7, shape=16, mapping=aes(color=location)) +
  coord_flip() +
  scale_color_loc() +
  scale_y_log10(name="Unique human-viral read fraction") +
  guides(color=guide_legend(nrow=2)) +
  theme_base + theme(axis.title.y = element_blank())
g_hv_box
```

The effect of panel enrichment on HV read fraction was much stronger than on total viral read fraction, with overall relative abundance increasing from 1.23e-05 to 1.14e-02, or almost 1000-fold.

## Overall taxonomy and composition

HV reads in enriched samples are dominated by coronaviruses and parvoviruses, which makes sense given the composition of the panel and what we saw for total-virus composition. Unenriched samples' HV composition, meanwhile, was quite unusual for wastewater: while some standard enteric virus groups (e.g. Sedoreoviridae, Picornaviridae) were still abundant here, a large fraction of HV reads were mapped to *Polyomaviridae* and *Papillomaviridae*, two groups that seldom show a large presence in wastewater data.

```{r}
#| label: raise-hv-taxa
#| edit-manually: false

# Filter samples and add viral taxa information
mrg_hv_filtered_named <- mrg_hv_filtered %>% left_join(viral_taxa, by="taxid") 

# Discover viral species & genera for HV reads
hv_reads_genus <- raise_rank(mrg_hv_filtered_named, viral_taxa, "genus")
hv_reads_family <- raise_rank(mrg_hv_filtered_named, viral_taxa, "family")
```

```{r}
#| label: fig-hv-family
#| fig-height: 6
#| fig-width: 6
#| edit-manually: false
#| fig-cap: "Family composition of putative human-viral reads across samples."


threshold_major_family <- 0.01

# Count reads for each human-viral family
hv_family_counts <- hv_reads_family %>% 
  group_by(name, taxid, across(all_of(grouping_cols))) %>%
  count(name = "n_reads_hv") %>%
  group_by(across(all_of(grouping_cols))) %>%
  mutate(p_reads_hv = n_reads_hv/sum(n_reads_hv))
hv_family_counts_collapsed <- hv_family_counts %>% 
  mutate(minor = p_reads_hv < threshold_major_family,
         name_display = ifelse(minor, "Other", name),
         taxid_display = ifelse(minor, NA, taxid)) %>%
  group_by(across(all_of(grouping_cols)), name_display, taxid_display) %>%
  summarize(n_reads_hv = sum(n_reads_hv), p_reads_hv = sum(p_reads_hv), .groups = "drop")
hv_family_levels <- hv_family_counts_collapsed %>% 
  group_by(is.na(taxid_display), name_display) %>% summarize(.groups = "drop") %>%
  pull(name_display)
hv_family_counts_display <- hv_family_counts_collapsed %>%
  rename(p_reads = p_reads_hv, label = name_display) %>%
  mutate(label = factor(label, levels = hv_family_levels))

# Get most prominent families for text
hv_family_counts_collate <- hv_family_counts %>%
  group_by(name, taxid, across(all_of(grouping_cols_agg))) %>%
  summarize(n_reads_tot = sum(n_reads_hv),
            p_reads_max = max(p_reads_hv), .groups = "drop") %>%
  arrange(desc(n_reads_tot))

# Plot
g_hv_family <- g_comp_base +
  geom_composition(data=hv_family_counts_display) +
  ggtitle("Family composition of human-viral reads") +
  scale_y_composition(name="% HV Reads") +
  scale_fill_viral(name="Viral family")
g_hv_family
```

Given my other research goals for this post, I'm not going to dig into this in detail here. But it might be worth a more detailed look in the future.

## Length and complexity metrics

Some of the most important changes in this version of the pipeline were additional measurements of fragment length and duplication levels in the HV read subset. In this section, I'll investigate these new metrics. This analysis largely follows Harmon's investigation of the same data [here](https://data.securebio.org/harmons-public-notebook/notebooks/2024-10-16-frag_dup_analysis.html).

### Fragment lengths

We measured fragment length in two ways: firstly, as the alignment length to the HV reference calculated by Bowtie2, and secondly as the length of the FASTA sequence following read pair merging with BBMerge. The latter method is only applicable to read pairs with enough overlap to be successfully merged with BBMerge; fragments too long for this to occur cannot be accurately measured this way. For the Crits-Christoph data, about 14% of HV sequences were successfully merged by BBMerge and thus have both length metrics available.

```{r}
#| label: fig-hv-fragment-length
#| fig-cap: "Kernel density plot of fragment lengths in Crits-Christoph data, measured using different methods."

# Prepare subdatasets
frag_bb <- mrg_hv_filtered %>%
  filter(bbmerge_merge_status == 1) %>%
  transmute(length = bbmerge_frag_length, group="BBMerge")
frag_bt_merged <- mrg_hv_filtered %>%
  filter(bbmerge_merge_status == 1) %>%
  transmute(length = bowtie2_frag_length, group="Bowtie2 (merged only)")
frag_bt_unmerged <- mrg_hv_filtered %>%
  filter(bbmerge_merge_status == 0) %>%
  transmute(length = bowtie2_frag_length, group="Bowtie2 (unmerged only)")
frag_bt_all <- mrg_hv_filtered %>%
  transmute(length = bowtie2_frag_length, group="Bowtie2 (all)")
tab_frag <- bind_rows(frag_bb, frag_bt_merged, frag_bt_unmerged, frag_bt_all)

# Plot
g_frag_density <- ggplot(tab_frag, aes(x=length, color=group)) +
  geom_density() +
  scale_color_brewer(palette="Set1", name="Method") +
  labs(y="Density", x="Fragment length") +
  theme_base
g_frag_density
```

```{r}
#| label: fig-frag-cor
#| fig-cap: "Scatterplot showing Bowtie2- and BBMerge-measured fragment lengths for the subset of reads that merged successfully with BBMerge. Red line indicates x=y; R gives the Spearman's rank correlation between the two variables."

g_frag_cor <- mrg_hv_filtered %>%
  filter(bbmerge_merge_status == 1) %>%
  ggplot(aes(x=bowtie2_frag_length, y=bbmerge_frag_length)) +
  geom_abline(color="red") +
  geom_point(shape=16, alpha=0.3) +
  stat_cor(method = "spearman", label.x.npc = 0.8) +
  labs(x="Fragment length (Bowtie2)", y="Fragment length (BBMerge)") +
  theme_base
g_frag_cor
```

-   For sequences that merged successfully, the Bowtie2 and BBMerge length metrics are very similar (Spearman's rho \> 0.8), indicating a high level of agreement between the two approaches.

-   Unsurprisingly, sequences that did not successfully merge with BBMerge show much longer Bowtie2-measured fragment lengths than those that did, consistent with the former arising from fragments that were too long to achieve sufficient overlap between the forward and reverse reads.

-   Across all HV reads, Bowtie2-measured fragment length (post-preprocessing) ranged from 36 to 549nt, with a mean of 214.5nt.

### Duplication

As with fragment length, we identified duplicates in the data via two different methods:

1.  Duplicates were identified using **Bowtie2** by looking for read pairs that aligned to the same reference genome with the same start and end coordinates (+/- 1 nucleotide position).
2.  Duplicates were identified using **Clumpify** (from the BBTools suite) by looking for read pairs with near-identical sequences, with a small amount of tolerance for sequencing errors.

Importantly, the Bowtie2 duplicate analysis identified duplicate groups but did not discard identified duplicates, while the Clumpify analysis both identified duplicate groups and performed deduplication. The latter deduplication also discarded some information about the output of the former method, which somewhat limits our ability to analyze and compare duplication patterns across the two methods. This situation isn't ideal, and is something we want to address in future pipeline versions.

Proceeding with the analysis given these limitations, we find that, given the information available, the two methods produce very similar distributions of duplication behavior:

```{r}
#| label: fig-density-dupcount
#| fig-cap: "Distribution of duplicate group sizes in the HV data, as measured by both Clumpify and Bowtie2."

# Calculate dupcount distribution
tab_dupcount_clumpify <- mrg_hv_filtered %>% count(clumpify_dupcount) %>%
  rename(dupcount = clumpify_dupcount) %>% mutate(method="Clumpify")
tab_dupcount_bowtie2  <- mrg_hv_filtered %>% group_by(bowtie2_exemplar) %>%
  slice_head(n=1) %>% ungroup %>% count(bowtie2_dupcount) %>%
  rename(dupcount = bowtie2_dupcount) %>% mutate(method="Bowtie2")
tab_dupcount <- bind_rows(tab_dupcount_clumpify, tab_dupcount_bowtie2) %>%
  group_by(method) %>% mutate(p=n/sum(n)) %>%
  mutate(reads_total = n * dupcount)

# Plot
g_dupcount_density <- ggplot(tab_dupcount, 
                             aes(x=dupcount, y=n, color=method)) +
  geom_point(shape=16, alpha=0.7) +
  scale_color_brewer(palette="Dark2", name="Method") +
  scale_y_log10() +
  labs(y="# Duplicate groups", x="Reads in duplicate group") +
  theme_base
g_dupcount_density
```

In both cases, the great majority of sequences (\>92%) are only represented by a single read, and the great majority of the remainder (\>80%) are represented by two or three reads. This broad pattern continues into progressively higher duplicate counts. The overall fraction of duplicate reads (i.e. reads that aren't the first in their duplicate group) was around 9% under both methods.

There is a moderate correlation between the two methods (r = 0.47) in terms of the size of each duplicate group:

```{r}
#| label: fig-dupcount-cor
#| fig-cap: "Scatterplot showing Bowtie2- and Clumpify-measured duplicate counts for each duplicate group. Red line indicates x=y; R gives the Pearson correlation between the two variables."

tab_dupcount_cor <- mrg_hv_filtered %>% group_by(bowtie2_exemplar) %>%
  slice_head(n=1)

g_dupcount_cor <- ggplot(tab_dupcount_cor,
                         aes(x=clumpify_dupcount, y=bowtie2_dupcount)) +
  geom_abline(color="red") +
  geom_point(shape=16, alpha=0.3) +
  stat_cor(method = "pearson", label.x.npc = 0.8) +
  scale_y_continuous(breaks = seq(0,40,4)) +
  labs(x="Duplicate group size (Clumpify)", y="Duplicate group size (Bowtie2)") +
  theme_base
g_dupcount_cor
```
